Models

### 2021-03-12 analysis/05_model_metrics.Rmd
Clean up modeling and visualization

metrics <- pins::pin_get("rarefied-medians", board = "github") %>%
  select(-cYEAR) %>%
  pivot_longer(., cols = setdiff(everything(), one_of("rarefyID", "STUDY_ID", "SamplePool", "SampleN", "num_years", "duration", "startYear", "endYear", "YEAR", "cell", "rarefied", "type", "realm", "climate", "habitat", "biome_map", "taxa", "organisms", "cent_lat", "cent_long", "abundance_type", "BROAD_TYPE", "taxa_mod", "climate_mod")), names_to = "metric", values_to = "value") %>%
  rename_with(tolower, .cols = setdiff(everything(), one_of("rarefyID")))  %>%
  mutate(logvalue = log(value + abs(floor(min(na.omit(value))))),
         year_scaled = scale(year)) %>% # create a column for log transformation that adds the absolute value integer closest to the most negative value to create all positive values
  filter(!is.na(value)) 
```
####Data exploration

@@

metric_mean_fit_slope <- metrics %>%
  filter(metric %in%  c("FRic", "FEve", "FDiv", "S_int")) %>%
  group_nest(metric) %>%
  mutate(model = map(data, ~lmer(logvalue ~ year_scaled + (year_scaled|study_id/rarefyID), 
                                 data = .x)),
           null_model = map(data, ~lmer(logvalue ~ (year_scaled|study_id/rarefyID), 
                                 data = .x)),
           test = map2(model, null_model, anova),
           test_res = map(test, tidy),
           coef = map(model, tidy)
           ) %>% 
  unnest(cols = c(test_res)) %>%
  select(metric, model, coef, p.value) %>%
  filter(!is.na(p.value)) %>%
  unnest(cols = c(coef)) %>%
  filter(term == "year_scaled") %>%
  select(-c(effect, group, term, statistic))
```

Fit model to individual studies - if there is more than one timeseries in a study, account for that
```{r}
sing_study <- metrics %>%
  group_by(study_id) %>%
  filter(n_distinct(rarefyID) == 1) %>% 
  pull(study_id) %>% unique()

# indiv_data <- metrics %>%
#   mutate(rarefyID = as.factor(rarefyID)) %>%
#   group_by_at(setdiff(names(metrics), c("year", "value", "logvalue", "rarefyID", "cell", "samplen"))) %>% 
#   #filter(n_distinct(year) > 1) %>% #filter out rarefyID's that don't have multiple years after dropping Na's
#   group_nest()

indiv_data <- metrics %>%
  mutate(rarefyID = as.factor(rarefyID)) %>%
  group_by(study_id) %>%
  group_nest()

indiv_ts_sing_fit <- indiv_data %>%
  filter(study_id %in% sing_study) %>%
  mutate(model = map(data, ~ lm(logvalue ~ year, data = .x))) %>%
  mutate(model_type = "lm")

indiv_ts_fit <- indiv_data %>%
  filter(!study_id %in% sing_study) %>%
  mutate(model = map(data, ~lmer(logvalue ~ (1 + year|rarefyID), 
                                 data = .x, control = lmerControl(optimizer ="Nelder_Mead")))) %>%
  mutate(model_type = "mixed",
         p_value = map(model, ~drop1(.x, test = "Chisq") %>% select(`Pr(Chi)`) %>% slice(2) %>% pull()))

indiv_ts_coef <- bind_rows(indiv_ts_fit, indiv_ts_sing_fit) %>%
  mutate(coef = map(model, tidy)) %>% 
  select(-data, -model) %>% 
  unnest(cols = c(coef))
```
Fit a single model to play around with 
```{r}
eve <- metrics %>% 
  filter(metric == "FEve")

feve_fit <- glmer(value ~ year + (1|study_id/rarefyID), data = eve, family = Gamma,
                  control = glmerControl(optCtrl = list(maxit = 1e9,
                                                    maxfun = 1e9,
                                                    xtol_abs = 1e-11,
                                                    ftol_abs = 1e-11,
                                                    maxeval = 1e9)))

#find best optimizer
diff_optims <- allFit(feve_fit, maxfun = 1e5)

optimx_options <- c("L-BFGS-B", "nlminb", "nlm", "bobyqa", "nmkb", "hjkb")

for(i in 1:length(optimx_options)){
  model_flex <- glmer(value ~ year + (1|study_id/rarefyID), data = eve, family = Gamma,
                     control = glmerControl(optimizer = "optimx",
                                           optCtrl = list(method = optimx_options[i],
                                                                   maxit = 1e9)))
  print(model_flex@optinfo$conv$lme4$messages)
}

algoptions <- c("NLOPT_LN_PRAXIS", "NLOPT_GN_CRS2_LM",
"NLOPT_LN_COBYLA", "NLOPT_LN_NEWUOA",
"NLOPT_LN_NEWUOA_BOUND", "NLOPT_LN_NELDERMEAD",
"NLOPT_LN_SBPLX", "NLOPT_LN_BOBYQA")

for(i in 1:length(algoptions)){
  model_flex <- glmer(value ~ year + (1|study_id/rarefyID), data = eve, family = Gamma,
                     control = glmerControl(optimizer = "nloptwrap",
                                           optCtrl = list(algorithm = algoptions[i],
                                                          maxeval = 1e7,
                                                          xtol_abs = 1e-9,
                                                          ftol_abs = 1e-9)))
  if(is.null(model_flex@optinfo$conv$lme4$messages)){
    print(paste0("One of the nloptwrap options, ", algoptions[i],", worked!"))
    print(summary(model_flex))
    break
  }
}

fit_mod <- function(data){
  print(unique(data$rarefyID)) %>%
  lmer(logvalue ~ (1 + year|rarefyID), 
                                 data = data, control = lmerControl(optimizer ="Nelder_Mead"))
}

indiv_data %>%
  filter(!study_id %in% sing_study) %>%
  mutate(model = map(data, ~lmer(logvalue ~ ( 1+ year|rarefyID), 
                                 data = .x, control = lmerControl(optimizer ="Nelder_Mead"))))

fit_mod <- function(filter_id){
  filter_data <- metrics %>% filter(study_id == filter_id)

  print(filter_id)
  lmer(logvalue ~ (1+ year|rarefyID), 
                                 data = filter_data)
}

mult_studies <- metrics %>% filter(!study_id %in% sing_study) %>% pull(study_id) %>% unique()
map(mult_studies, fit_mod)

Plot of S and diversity metrics
```{r}
indiv_ts_coef %>%
  filter(term == "year", metric %in% c("FRic", "FEve", "FDiv", "FDis", "S")) %>%
  select(metric, estimate) %>% 
  ggplot() +
  geom_histogram(aes(x = estimate)) + #,
                 #binwidth = 2) +
  scale_x_continuous(name = 'slope') +
  #scale_y_continuous(breaks = c(0,2500,5000,7500,10000,12500, 15000, 17500, 20000,22500)) +
  labs(tag = 'B',
       y = 'Number of cells') +
  theme_bw() +
  theme(panel.grid.minor.x = element_blank()) +
  facet_wrap(vars(metric), scales = "free")
```
Reproduce Dornelas figure - all data points & regressions w/overall mean
```{r}
#get climate classification
metadata <- read_csv(system.file("extdata", "biotime/biotime_metadata.csv", package = "biodivTS"))

@@

plot_metric <- function(metric_name, ylabel){

  data <- metrics %>%
    filter(metric == metric_name, !is.na(value), !is.infinite(value))

  model_coef <- metric_mean_coef %>%
    filter(metric == metric_name) %>%
    select(term, estimate)

  intercept <- model_coef %>% filter(term == "(Intercept)") %>% pull(estimate)
  slope <- model_coef %>% filter(term == "year") %>% pull(estimate)

  pred <- data.frame(year = min(data$year):max(data$year)) %>%
    mutate(pred = intercept + year * slope)

  trend_plot <- data %>%
    left_join(pred, by = "year") %>%
    #left_join(metadata %>% select(study_id, climate)) %>%
    ggplot(aes(x = year, y = logvalue)) +
    # geom_smooth(aes(color = climate, group = rarefyID), method = "glm", se = FALSE) +
    # geom_point(aes(color = climate, group = rarefyID)) +
    geom_point(aes(group = rarefyID) ,color = "grey", size = 0.30) +
    geom_smooth(aes(color = climate_mod, group = rarefyID), method = "glm", se = FALSE, size = 0.5) +
    #geom_smooth(color = "black", method = "glm", se = FALSE) +
    geom_line(mapping = aes(y = pred)) +
    theme(legend.position = "none") +
    theme_classic() +
    ylab(ylabel) +
    labs(color = "Climate")

@@

@@

plot_map <- tibble(metric_name = c("S", "FRic", "FEve", "FDiv", "FDis"), 
           ylabel = c("log(S)", "log(FD)", "log(Evenness)", "log(Divergence)", "log(Dispersion)"))
pmap(plot_map, plot_metric)
```

>>>>

metrics <- pins::pin_get("rarefied-medians", board = "github") %>%
  pivot_longer(., cols = setdiff(everything(), one_of("rarefyID", "STUDY_ID", "SamplePool", "SampleN", "num_years", "duration", "startYear", "endYear", "YEAR", "cell", "rarefied", "type", "realm", "climate", "habitat", "biome_map", "taxa", "organisms", "cent_lat", "cent_long", "abundance_type", "BROAD_TYPE", "taxa_mod", "climate_mod")), names_to = "metric", values_to = "value") %>%
  rename_with(tolower, .cols = setdiff(everything(), one_of("rarefyID")))  %>%
  mutate(logvalue = log(value + abs(floor(min(na.omit(value))))),
         year_scaled = scale(year)) %>% # create a column for log transformation that adds the absolute value integer closest to the most negative value to create all positive values
  filter(!is.na(value)) 
```
####Data exploration

@@

metric_mean_fit_slope <- metrics %>%
  filter(metric %in%  c("FRic", "FEve", "FDiv", "S_int")) %>%
  group_nest(metric) %>%
  mutate(model = map(data, ~lmer(logvalue ~ year_scaled + (year_scaled|study_id/rarefyID), 
                                 data = .x)),
           null_model = map(data, ~lmer(logvalue ~ (year_scaled|study_id/rarefyID), 
                                 data = .x)),
           test = map2(model, null_model, anova),
           test_res = map(test, tidy),
           coef = map(model, tidy)
           ) %>% 
  unnest(cols = c(test_res)) %>%
  select(metric, model, coef, p.value) %>%
  filter(!is.na(p.value)) %>%
  unnest(cols = c(coef)) %>%
  filter(term == "year_scaled") %>%
  select(-c(effect, group, term, statistic))

model_comp <- metric_mean_fit %>%
  select(metric, no_slope_estimate = estimate, no_slope_pval = p.value) %>%
  left_join(metric_mean_fit_slope %>% 
  select(metric, slope_estimate = estimate, slope_pval = p.value ))
```

Reproduce Dornelas figure - all data points & regressions w/overall mean
```{r}
#get climate classification
metadata <- read_csv(system.file("extdata", "biotime/biotime_metadata.csv", package = "biodivTS"))

@@

plot_metric <- function(metric_name, ylabel, data, model_coef){

  data <- data %>%
    filter(metric == metric_name, !is.na(value), !is.infinite(value))

  model_coef <- model_coef %>%
    filter(metric == metric_name) %>%
    select(term, estimate)

  intercept <- model_coef %>% filter(term == "(Intercept)") %>% pull(estimate)
  slope <- model_coef %>% filter(term == "year_scaled") %>% pull(estimate)

  pred <- data %>%
    select(year_scaled) %>%
    distinct() %>%
    mutate(pred = intercept + year_scaled * slope)

  trend_plot <- data %>%
    left_join(pred, by = "year_scaled") %>%
    #left_join(metadata %>% select(study_id, climate)) %>%
    ggplot(aes(x = year_scaled, y = logvalue)) +
    # geom_smooth(aes(color = climate, group = rarefyID), method = "glm", se = FALSE) +
    # geom_point(aes(color = climate, group = rarefyID)) +
    geom_point(aes(group = rarefyID) ,color = "grey", size = 0.30) +
    geom_smooth(aes(color = climate_mod, group = rarefyID), method = "glm", se = FALSE, size = 0.5) +
    #geom_smooth(color = "black", method = "glm", se = FALSE) +
    geom_line(mapping = aes(y = pred)) +
    theme(legend.position = "none") +
    theme_classic() +
    ylab(ylabel) +
    labs(color = "Climate")

@@

model_coef <- metric_mean_fit_slope %>%
  select(metric, model) %>%
   mutate(coef = map(model, tidy)) %>% 
  select(-model) %>% 
  unnest(cols = c(coef))

@@

plot_map <- tibble(metric_name = c("S_int", "FRic", "FEve", "FDiv", "FDis"), 
           ylabel = c("log(S_int)", "log(FD)", "log(Evenness)", "log(Divergence)", "log(Dispersion)"))
pmap(plot_map, plot_metric, data = metrics, model_coef = model_coef)
```
